DRAFT: Harmonisation and trait aggregation paper

### 2020-12-09 Paper/Draft_harmonization_paper.tex
09.12: Final edits before send to Ralf

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{DRAFT: Harmonized and trait aggregation paper }
\author{}%Stefan Kunz 
\date{}

@@

To our knowledge, only Brown et al. \cite{brown_functional_2018} harmonized grouping features from more than two geographically distant invertebrate trait databases, for a limited set of grouping features (8) and taxa (112), in a study on the influence of decreasing glacier cover on functional diversity and community assembly of invertebrates. We suspect that the heterogeneity of information in freshwater invertebrate trait databases, besides the diversity of taxa across regions, is likely a major reason for the lack of studies across continents. To harmonize grouping features from different regions, first commonly accepted and unambiguous trait definitions are required \cite{schneider_towards_2019}. In the best case, grouping features would be classified into the same traits across databases or they could easily be harmonized using standardized terminology. However, a lack of standardized terminology of trait definitions and poor metadata quality in many trait databases are common issues throughout the field of trait-based ecology \cite{baird_toward_2011, kissling_towards_2018}. Secondly, consistent coding of traits facilitates the compatibility of trait data from different databases.

@@

However, invertebrate trait databases are heterogeneous regarding the coding they use for their traits \cite{culp_incorporating_2011}. For example, in the study from Brown et al. \cite{brown_functional_2018} the authors needed to reclassify traits using expert knowledge given that the individual databases employed different coding approaches (i.e. European and New Zealand databases fuzzy coding, North American database binary coding). A centralized database with standardized and unambiguous traits and a consistent coding of traits would minimize the data processing effort. Discrepancies in the taxonomic resolutions (e.g. species, genus or family) when linking observational taxonomic data to trait database represents another challenge. When observations are on a more precise taxonomic level than data available in the trait databases (e.g. observations on species-level, trait data on genus-level) trait information of the less precise taxonomic level is often assigned (e.g. \cite{szocs_effects_2014, vos_taxonomic_2017}). Conversely, if trait information is only available on more precise taxonomic levels than the observed taxa, traits are aggregated to a less precise taxonomic level, e.g. \cite{poff_functional_2006, szocs_effects_2014, piliere_a._f._h._importance_2016, aspin_extreme_2019}. Studies have used different methods of trait aggregation, e.g. the mean \cite{magliozzi_functional_2019}, median \cite{szocs_effects_2014} or the mode \cite{piliere_a._f._h._importance_2016}. Up to now, studies on how and to which extent different trait aggregation methods influence trait-based analysis are missing. However, related knowledge would inform future studies regarding the choice of the aggregation method.

The outlined difficulties when working with or synthesizing trait data motivated us to (1) investigate discrepancies in trait definition between trait databases from Europe, North America, Australia, and New Zealand. As a basis for further evaluating different trait aggregation methods and how harmonising definition discrepancies between grouping features can influence trait-environment relationships, we established 4 novel invertebrate trait datasets from these trait databases by harmonising 7 grouping features. We evaluate different trait aggregation methods by (2) comparing aggregated traits with assigned traits by experts at family-level for two of the established trait datasets. (3) we re-analysed data on the effect of anthropogenic salinisation on biological traits by Szöcs et al. \cite{szocs_effects_2014} using trait data from the established harmonized European trait datasets and aggregated traits. By comparison with the original analysis, we investigated how harmonizing and aggregating trait data can modify the outcome of trait-environment relationships.

@@

We extracted information from 6 trait databases from Europe, North America, Australia, and New Zealand and harmonized 7 grouping features. Trait information for Europe was obtained from the freshwaterecology.info database \cite{schmidt-kloiber_www.freshwaterecology.info_2015} and the Tachet database \cite{usseglio-polatera_biomonitoring_2000}. The freshwaterecology.info contains taxa on species-level, while taxa recorded in the Tachet database are on species, genus and family-level. Trait information for North America was obtained from Twardochleb et al. \cite{twardochleb_trait_data_2020} and complemented by Vieira et al. \cite{vieira_database_nodate}. Data on body form for European and North American taxa was based on expert knowledge \cite{polatera_personal_information_2020}. For Australia and New Zealand, we used trait databases from Kefford et al. \cite{kefford_integrated_2020} and Philips and Smith respectively \cite{Philips_and_Smith_NZ_DB_2018}. To increase readability we refer to the databases as well as the datasets we extracted from them by the name of the continent they originate from, except for European databases, which we refer to by their commonly used names (freshwaterecology.info and Tachet database).

We selected traits of seven grouping features that were available in all databases, are commonly applied in trait-based ecological studies, and describe different parts of the biology of a species: life history (Voltinism), morphology (Respiration, Body form, Size), ecology (Locomotion, Feeding mode) and reproduction (Oviposition). We omitted ecological traits that describe habitat preferences (e.g. temperature preference) because these traits are missing in the New Zealand trait database. The grouping features were differently classified across the databases, we therefore harmonized them into 26 traits (Table \ref{tab:traits_harmonization}). Harmonization was undertaken by amalgamating similar traits into one trait (e.g. crawlers and sprawlers into crawlers). Thereby, for a particular taxa the highest trait affinity score among the amalgamated traits was taken.

We used fuzzy coding for establishing our harmonized datasets unless data quality prohibited. In the latter case we used binary traits, i.e. categorical and continuous traits were converted into binary traits. For example, in the  freshwaterecology.info database, the classification of the trait voltinism accounts for different faunistic regions. Hence, entries such as "arctic" or "boreal" in the voltinism traits were substituted with a value of 1. Implicitly, we assumed for binary traits that a value of 1 and 0 corresponds to the highest and no affinity for a particular trait. Fuzzy codes are reported with different ranges in the trait databases (e.g. freshwaterecology.info 0 to 10, Tachet database 0 to 3 or 0 to 5). We standardized these to a range between 0 and 1 and converted trait affinities to percentages. Thus, fuzzy coded and binary traits were in the same range.

Prior to harmonization we consolidated duplicate taxa on species, genus or family-level if present within a dataset, by either applying the median for fuzzy coded traits, or the maximum for binary traits. We omitted taxa with a lower taxonomic precision than family-level.

@@

\caption{Traits of the harmonized grouping features. The last column indicates traits that were amalgamated for harmonization (no amalgamation needed if empty).}

@@

Traits of the harmonized grouping feature datasets were aggregated to family-level using three approaches. I) Direct aggregation of taxa to family-level giving equal weight to every taxon using the mean or median, denoted \textit{direct\_agg\textsubscript{mean}} and \textit{direct\_agg\textsubscript{median}}, respectively. II) Stepwise aggregation, i.e. first to the genus-level and subsequently to the family-level using the mean or median. This approach gives equal weights to each genus. Hereafter, we denote this aggregation type as \textit{stepwise\_agg\textsubscript{mean}} or \textit{stepwise\_agg\textsubscript{median}}, respectively. III) Aggregation using a weighted mean approach, denoted as \textit{weighted\_agg}. This method weights the genera according to the number of their species in the trait datasets regardless if for every used grouping feature information was available (Figure \ref{fig:data_proc_overview}).

To examine the influence of the taxonomic hierarchy and the trait variability on the outcome of the different trait aggregation methods we created three hypothetical examples of different taxonomic hierarchies (scenarios). 
1) A family with an equal number of genera and species (5 genera each with 5 species respectively), denoted as \textit{sim\_base}.
2) A family where one genus has a much larger number of species than the other 4 genera (1 genus with 13 species, 4 genera with 3 species respectively), denoted as \textit{sim\_extreme}. 
3) A family where all genera have a different number of species (8, 2, 7, 3, 5), denoted as \textit{sim\_variation}. In total every family consisted of 25 species. To each hypothetical taxonomic hierarchy a hypothetical grouping feature with 3 traits was assigned. The 25 affinities for each trait were simulated by sampling from a truncated normal distribution with a mean value of 0.5 and 5 levels of standard deviations (0.2, 0.4, 0.6, 0.8, and 1) respectively to simulate different levels of trait variability. The truncated normal distribution was bound to 0 and 1. Simulated trait affinities were converted to percentages, similar to the data processing of the trait databases. The sampling was repeated 100 times for each standard deviation. Hence, in total 100 datasets for each level of trait variability were simulated, resulting in 12500 simulated trait affinities for each scenario ($25$ species per family $* 5$ levels of trait variability $* 100$ replicates). All trait aggregation methods were applied to each simulated dataset resulting in 500 aggregated trait values per scenario. The results were compared based on the produced range of aggregated trait affinities between levels of trait variability and hypothetical taxonomic hierarchies. We also compared for each simulated dataset the differences in trait affinities obtained by each aggregation method.

@@

We repeated the analysis of Szöcs et al. \cite{szocs_effects_2014} who studied the effect of anthropogenic salinization on invertebrates in the River Werra in Germany. For the re-analysis we used the established harmonized grouping features for Europe and additionally aggregated traits using the aforementioned aggregation methods.

@@

We substituted 6 of the grouping features from the original data with harmonized grouping features from the European trait dataset. We compared the results of the redundancy analysis (RDA) from the original study to the case when using harmonized grouping features. Specifically, the trait composition expressed as community weighted mean (CWM) traits was ordinated along an electric conductivity gradient. We compared the species scores obtained from the RDA, i.e. the coordinates of the tips of the vectors representing the CWM traits in the bi- or triplots. Following the original study, we identified traits associated with high or low salinity based on their distance to the ordination axis median using the mahalanobis distance. Traits with a mahalanobis distance greater than the $97.5 \%$ -quantile of the Chi-square distribution (5.02) were regarded as traits responding to either low or high salinity. For our analysis, we used the same 21 grouping features that Szöcs et al. \cite{szocs_effects_2014} used. The 6 harmonized grouping features used were \textit{Size, Feeding mode, Locomotion, Oviposition, Respiration}, and \textit{Voltinism}. Additionally, for testing the effect of aggregated traits we assigned to each taxon in Szöcs et al. \cite{szocs_effects_2014} the aggregated trait value for its corresponding family and repeated the RDA.

@@

\subsection*{Taxonomic coverage of the harmonized trait datasets}

@@

\begin{table}[ht]
    \centering
    \caption{Number of taxa per harmonized dataset and per taxonomic level. Numbers in parenthesis show rounded relative frequencies in percentage.} 
    \label{tab:tax_coverage}
    \begin{tabular}{lccccc}
    \toprule[.1em]
    Database & Nr. of taxa & Species & Genus & Family & Nr. aquatic insects \\ 
    \toprule[.1em]
    EU & 4601 & 3739 (81) & 704 (15) & 158 (3) & 3942 (86) \\ 
    NOA & 3753 & 2414 (64) & 1163 (31) & 176 (5) & 3305 (88) \\ 
    AUS & 1402 & 564 (40) & 578 (41) & 260 (19) & 1016 (72) \\ 
    NZ & 478 & 404 (85) & 47 (10) & 27 (6) & 443 (93) \\ 
    \bottomrule
    \end{tabular}
\end{table}

@@

The amount of entries with available information for the selected grouping features varied strongly for the preprocessed European, North American, and Australian datasets (Table \ref{tab:trait_coverage}). By contrast, the New Zealand dataset contained complete trait information for most of the investigated grouping features (between 94 \% and 100 \%).

@@

Comparing the aggregated trait affinities, in total 15000 ($500$ aggregated traits $*10$ unique comparisons of trait affinities and $3$ scenarios) for each simulated dataset showed that in most simulated datasets the differences between the aggregated traits were small. Only 1.42 \% (213 out of 15.000) of all comparisons showed a difference equal or greater than an absolute trait affinity of 0.1. Most of these differences occurred for the \textit{sim\_extreme} scenario (83.5 \%). The majority of the differences equal or above 0.1 were found between the aggregation methods \textit{direct\_agg\textsubscript{mean}} and \textit{stepwise\_agg\textsubscript{median}}, \textit{direct\_agg\textsubscript{median}} and \textit{stepwise\_agg\textsubscript{median}}, and \textit{stepwise\_agg\textsubscript{median}} and \textit{weighted\_agg} (Figure \ref{fig:sim_indv_runs}). 
% Differences were rather small, largest difference 0.265 for sim_extreme (stepwise_median vs weighted_agg)

Overall, differences between the results of the aggregation methods for each simulated dataset increased with increasing trait variability. According to the simulation results, the different aggregation methods often result in similar trait affinities.  If differences occurred, than mostly between aggregation methods using the median and aggregation methods using the mean.

@@

\subsection*{Re-analysis of Szöcs et al. using harmonized and aggregated grouping features}

@@

Using harmonized grouping features resulted in less traits that distinguish upstream and downstream sites in comparison to the original analysis (Figure \ref{fig:rda_species_scores} and Figure \ref{fig:boxplots_scores_on_constrained_axis}). According to the RDA of the trait composition, downstream sites were characterised by taxa with the traits multivoltinism, ovoviviparity and long life cycle duration. Upstream sites were characterised by univoltine taxa with a short life cycle duration that lay their eggs in an aquatic environment (aquatic eggs). The trait aquatic eggs of the harmonized grouping feature oviposition has been derived by amalgamating the trait oviposition in clutches and other related traits (Table \ref{tab:traits_harmonization}). The traits shredder and gill respiration did not characterise sites with high salinisation.

Using at family-level aggregated traits from the harmonized dataset showed also results similar to the original analysis but with fewer traits distinguishing upstream and downstream sites (Figure \ref{fig:rda_species_scores} and Figure \ref{fig:boxplots_scores_on_constrained_axis}). The \textit{direct\_agg\textsubscript{mean}}, \textit{direct\_agg\textsubscript{median}}, and \textit{weighted\_agg} 
characterised the downstream sites with the same traits as the original analysis except that downstream sites were not characterised by the trait shredder. Upstream sites were characterised by the traits univoltinism, aquatic eggs and short life cycle duration. The same results were obtained when re-analysing the data with traits aggregated by the \textit{stepwise\_agg\textsubscript{mean}} and \textit{stepwise\_agg\textsubscript{median}}, with the exception that none of the life cycle traits characterised upstream or downstream sites. Thus, in our comparison aggregating by \textit{direct\_agg\textsubscript{mean}}, \textit{direct\_agg\textsubscript{median}}, and \textit{weighted\_agg} yielded to the least change in interpretation of the RDA results compared to the original results.

\begin{figure}[H]
    \label{fig:rda_species_scores}
    \centering
    \includegraphics[width=16.5cm, height=10cm]{Species_scores_rda.png}
    \caption{Species scores obtained by RDA from the original analysis \cite{szocs_effects_2014}, using harmonized grouping features, and using harmonized grouping features with traits aggregated to family-level. The dots represent the individual species scores for each analysed trait along the conductivity axis. The violin plot shows the density estimate of the species scores. Gray vertical lines indicate the median of the obtained species scores.}
\end{figure}

@@

% Why do we need trait-based analysis, especially across regions?
Synthesizing trait information from multiple trait databases is crucial to unfold the full potential of trait-based approaches, e.g. for studying community trait responses to environmental gradients worldwide. Our attempt of harmonising invertebrate trait information from trait databases of different regions showed that grouping feature harmonisation is labour-intensive because grouping features are differentiated differently into traits between trait databases and various codings are used to describe traits. In addition, the same traits are sometimes defined differently in the investigated databases, requiring expert knowledge to reclassify these traits (e.g. the trait piercer in the Tachet database, Table \ref{tab:traits_harmonization}). Although, others have noted the lack of standardized trait terminology in freshwater ecology \cite{baird_toward_2011, brink_traits-based_2011} we are the first to describe invertebrate trait definition discrepancies for some commonly used grouping features. To resolve definition discrepancies and facilitate data synthesis in the future terminological standards are needed. Harmonised definitions and concepts of traits have been developed in the past for other organisms, e.g. for plants with the \textit{Thesaurus Of Plant characteristics} (TOP) initiative \cite{garnier_towards_2017}. The core of this harmonisation initiative is to provide standardized trait definitions and to draw connections to synonyms, related terms, and surrounding concepts by linking to other controlled vocabularies or ontologies in the field. Such initiatives could be a role model for freshwater ecologists to establish unambiguous terminologies for invertebrate traits. The existing freshwater invertebrate trait databases could be linked through standardised terminologies or ontologies, as suggested by Baird et al. \cite{baird_toward_2011} almost a decade ago. By following the recently proposed \textit{Ecological Trait-data Standard Vocabulary} (ETS) providers of invertebrate trait data could connect their traits to such ontologies \cite{schneider_towards_2019}. Once standardised trait definitions are established, these will improve trait data sharing, trait data processing when working with multiple trait databases, and ultimately interpretation of the derived results. % What is not mentioned here is how traits are measured, for which measurement protocols exist I believe. I think this is an separate issue.

@@

\subsection*{Trait aggregation and grouping feature harmonisation}

@@

% What do we really know because someone has watched the animals and published their behaviour; what is extrapolated knowledge? 
% Quality evaluation of traits als Thema..
Although comprehensive invertebrate trait databases have been developed for several regions in the past, data synthesis is difficult due to discrepancies in trait definitions. By providing an overview of definition discrepancies we hope to set a starting point for the development of standardized trait definitions through which trait databases can be linked. Standardized terminology will facilitate comparable trait analysis across large scales. 
% - Variation in taxonomic coverage 
We showed that trait aggregation is often in conclusion with expert assignments and specifically approaches using the median. ... weighting trait information based on the taxa present in the trait databases exerts a minor influence on the aggregated trait affinities. Which method the objectively better measure is cannot be answered, but when trait aggregation is necessary we showed 1) that highly variable traits should be treated carefully, 2) coding of traits should be considered, and 3) taxa where a few genera have a disproportional high number of species trait information .... If researchers need to aggregate traits, a measure of trait dispersion should be reported to indicate the uncertainty of the aggregated estimate. % Question for future research?
% Fuzzy coding and uncertainty?
% Efforts in creating trait ontologies etc etc
% Technology improvement, e.g. Deep learning for species identification

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{DRAFT: Harmonisation and trait aggregation paper }
\author{}%Stefan Kunz 
\date{}

@@

To our knowledge, only Brown et al. \cite{brown_functional_2018} harmonised grouping features from more than two geographically distant invertebrate trait databases, for a limited set of grouping features (8) and taxa (112), in a study on the influence of decreasing glacier cover on functional diversity and community assembly of invertebrates. We suspect that the heterogeneity of information in freshwater invertebrate trait databases, besides the diversity of taxa across regions, is likely a major reason for the lack of studies across continents. To harmonize grouping features from different regions, first commonly accepted and unambiguous trait definitions are required \cite{schneider_towards_2019}. In the best case, grouping features would be classified into the same traits across databases or they could easily be harmonised using standardised terminology. However, a lack of standardised terminology of trait definitions and poor metadata quality in many trait databases are common issues throughout the field of trait-based ecology \cite{baird_toward_2011, kissling_towards_2018}. Secondly, consistent coding of traits facilitates the compatibility of trait data from different databases.

@@

However, invertebrate trait databases are heterogeneous regarding the coding they use for their traits \cite{culp_incorporating_2011}. For example, in the study from Brown et al. \cite{brown_functional_2018} the authors needed to reclassify traits using expert knowledge given that the individual databases employed different coding approaches (i.e. European and New Zealand databases fuzzy coding, North American database binary coding). Databases with standardised and unambiguous traits and a consistent coding of traits would minimize the data processing effort. Discrepancies in the taxonomic resolutions (e.g. species, genus or family) when linking observational taxonomic data to trait database represents another challenge. When observations are on a more precise taxonomic level than data available in the trait databases (e.g. observations on species-level, trait data on genus-level) trait information of the less precise taxonomic level is often assigned (e.g. \cite{szocs_effects_2014, vos_taxonomic_2017}). Conversely, if trait information is only available on more precise taxonomic levels than the observed taxa, traits are aggregated to a less precise taxonomic level, e.g. \cite{poff_functional_2006, szocs_effects_2014, piliere_a._f._h._importance_2016, aspin_extreme_2019}. Studies have used different methods of trait aggregation, e.g. the mean \cite{magliozzi_functional_2019}, median \cite{szocs_effects_2014} or the mode \cite{piliere_a._f._h._importance_2016}. Up to now, studies on how and to which extent different trait aggregation methods influence trait-based analysis are missing. However, related knowledge would inform future studies regarding the choice of the aggregation method.

The outlined difficulties when working with or synthesizing trait data motivated us to (1) investigate discrepancies in trait definition between trait databases from Europe, North America, Australia, and New Zealand. As a basis for further evaluating different trait aggregation methods and how harmonising definition discrepancies between grouping features can influence trait-environment relationships, we established 4 novel invertebrate trait datasets from these trait databases by harmonising 7 grouping features. We evaluate different trait aggregation methods by (2) comparing aggregated traits with assigned traits by experts at family-level for two of the established trait datasets. (3) we re-analysed data on the effect of anthropogenic salinisation on biological traits by Szöcs et al. \cite{szocs_effects_2014} using trait data from the established harmonised European trait datasets and aggregated traits. By comparison with the original analysis, we investigated how harmonising and aggregating trait data can modify the outcome of trait-environment relationships.

@@

We extracted information from 6 trait databases from Europe, North America, Australia, and New Zealand and harmonised 7 grouping features. Trait information for Europe was obtained from the freshwaterecology.info database \cite{schmidt-kloiber_www.freshwaterecology.info_2015} and the Tachet database \cite{usseglio-polatera_biomonitoring_2000}. The freshwaterecology.info contains taxa on species-level, while taxa recorded in the Tachet database are on species, genus and family-level. Trait information for North America was obtained from Twardochleb et al. \cite{twardochleb_trait_data_2020} and complemented by Vieira et al. \cite{vieira_database_nodate}. Data on body form for European and North American taxa was based on expert knowledge \cite{polatera_personal_information_2020}. For Australia and New Zealand, we used trait databases from Kefford et al. \cite{kefford_integrated_2020} and Philips and Smith respectively \cite{Philips_and_Smith_NZ_DB_2018}. To increase readability we refer to the databases as well as the datasets we extracted from them by the name of the continent they originate from, except for European databases, which we refer to by their commonly used names (freshwaterecology.info and Tachet database).

We selected traits of seven grouping features that were available in all databases, are commonly applied in trait-based ecological studies, and describe different parts of the biology of a species: life history (Voltinism), morphology (Respiration, Body form, Size), ecology (Locomotion, Feeding mode) and reproduction (Oviposition). We omitted ecological traits that describe habitat preferences (e.g. temperature preference) because these traits are missing in the New Zealand trait database. The grouping features were differently classified across the databases, we therefore harmonised them into 26 traits (Table \ref{tab:traits_harmonization}). Harmonisation was undertaken by amalgamating similar traits into one trait (e.g. crawlers and sprawlers into crawlers). Thereby, for a particular taxa the highest trait affinity score among the amalgamated traits was taken.

We used fuzzy coding for establishing our harmonised datasets unless data quality prohibited. In the latter case we used binary traits, i.e. categorical and continuous traits were converted into binary traits. For example, in the  freshwaterecology.info database, the classification of the trait voltinism accounts for different faunistic regions. Hence, entries such as "arctic" or "boreal" in the voltinism traits were substituted with a value of 1. Implicitly, we assumed for binary traits that a value of 1 and 0 corresponds to the highest and no affinity for a particular trait. Fuzzy codes are reported with different ranges in the trait databases (e.g. freshwaterecology.info 0 to 10, Tachet database 0 to 3 or 0 to 5). We standardised these to a range between 0 and 1 and converted trait affinities to percentages. Thus, fuzzy coded and binary traits were in the same range.

Prior to harmonisation we consolidated duplicate taxa on species, genus or family-level if present within a dataset, by either applying the median for fuzzy coded traits, or the maximum for binary traits. We omitted taxa with a lower taxonomic precision than family-level.

@@

\caption{Traits of the harmonised grouping features. The last column indicates traits that were amalgamated for harmonisation (no amalgamation needed if empty).}

@@

Traits of the harmonised grouping feature datasets were aggregated to family-level using three approaches. I) Direct aggregation of taxa to family-level giving equal weight to every taxon using the mean or median, denoted \textit{direct\_agg\textsubscript{mean}} and \textit{direct\_agg\textsubscript{median}}, respectively. II) Stepwise aggregation, i.e. first to the genus-level and subsequently to the family-level using the mean or median. This approach gives equal weights to each genus. Hereafter, we denote this aggregation type as \textit{stepwise\_agg\textsubscript{mean}} or \textit{stepwise\_agg\textsubscript{median}}, respectively. III) Aggregation using a weighted mean approach, denoted as \textit{weighted\_agg}. This method weights the genera according to the number of their species in the trait datasets regardless if for every used grouping feature information was available (Figure \ref{fig:data_proc_overview}).

To examine the influence of the taxonomic hierarchy and the trait variability on the outcome of the different trait aggregation methods we created three hypothetical examples of different taxonomic hierarchies (scenarios). 
1) A family with an equal number of genera and species (5 genera each with 5 species respectively), denoted as \textit{sim\_base}.
2) A family where one genus has a much larger number of species than the other 4 genera (1 genus with 13 species, 4 genera with 3 species respectively), denoted as \textit{sim\_extreme}. 
3) A family where all genera have a different number of species (8, 2, 7, 3, 5), denoted as \textit{sim\_variation}. In total every family consisted of 25 species. To each hypothetical taxonomic hierarchy a hypothetical grouping feature with 3 traits was assigned. The 25 affinities for each trait were simulated by sampling from a truncated normal distribution with a mean value of 0.5 and 5 levels of standard deviations (0.2, 0.4, 0.6, 0.8, and 1) respectively to simulate different levels of trait variability. The truncated normal distribution was bound to 0 and 1. Simulated trait affinities were converted to percentages, similar to the data processing of the trait databases. The sampling was repeated 100 times for each standard deviation. Hence, in total 100 datasets for each level of trait variability were simulated, resulting in 12500 simulated trait affinities for each simulated trait ($25$ species per family $* 5$ levels of trait variability $* 100$ replicates). These 12500 simulated affinities per trait were assigned to each scenario. The above mentioned trait aggregation methods were applied to each simulated dataset. 
% ...resulting in 2500 aggregated affinities per trait and scenario ($12500$ trait affinities per trait divided by $25$, the number of species and multiplied by $5$, the number of aggregation methods). 
The results were compared based on the produced range of aggregated trait affinities between levels of trait variability and scenario. We also compared for each simulated dataset the differences in trait affinities obtained by each aggregation method.

@@

We repeated the analysis of Szöcs et al. \cite{szocs_effects_2014} who studied the effect of anthropogenic salinization on invertebrates in the River Werra in Germany. For the re-analysis we used the established harmonised grouping features for Europe and additionally aggregated traits using the aforementioned aggregation methods.

@@

We substituted 6 of the grouping features from the original data with harmonised grouping features from the European trait dataset. We compared the results of the redundancy analysis (RDA) from the original study to the case when using harmonised grouping features. Specifically, the trait composition expressed as community weighted mean (CWM) traits was ordinated along an electric conductivity gradient. We compared the species scores obtained from the RDA, i.e. the coordinates of the tips of the vectors representing the CWM traits in the bi- or triplots. Following the original study, we identified traits associated with high or low salinity based on their distance to the ordination axis median using the mahalanobis distance. Traits with a mahalanobis distance greater than the $97.5 \%$ -quantile of the Chi-square distribution (5.02) were regarded as traits responding to either low or high salinity. For our analysis, we used the same 21 grouping features that Szöcs et al. \cite{szocs_effects_2014} used. The 6 harmonised grouping features used were \textit{Size, Feeding mode, Locomotion, Oviposition, Respiration}, and \textit{Voltinism}. Additionally, for testing the effect of aggregated traits we assigned to each taxon in Szöcs et al. \cite{szocs_effects_2014} the aggregated trait value for its corresponding family and repeated the RDA.

@@

\subsection*{Taxonomic coverage of the harmonised trait datasets}

@@

\begin{table}[ht]
    \centering
    \caption{Number of taxa per harmonised dataset and per taxonomic level. Numbers in parenthesis show rounded relative frequencies in percentage.} 
    \label{tab:tax_coverage}
    \begin{tabular}{lccccc}
    \toprule[.1em]
    Database & Nr. of taxa & Species & Genus & Family & Nr. aquatic insects \\ 
    \toprule[.1em]
    EU & 4601 & 3739 (81) & 704 (15) & 158 (3) & 3942 (86) \\ 
    NOA & 3753 & 2414 (64) & 1163 (31) & 176 (5) & 3305 (88) \\ 
    AUS & 1402 & 564 (40) & 578 (41) & 260 (19) & 1016 (72) \\ 
    NZ & 478 & 404 (85) & 47 (10) & 27 (6) & 443 (93) \\ 
    \bottomrule
    \end{tabular}
\end{table}

@@

The amount of entries with available information for the selected grouping features varied strongly for the established European, North American, and Australian trait datasets (Table \ref{tab:trait_coverage}). By contrast, the New Zealand dataset contained complete trait information for most of the investigated grouping features (between 94 \% and 100 \%).

@@

Each trait aggregation method yielded one aggregated affinity for every simulated dataset, in total 15000 ($100$ replicates $*$ 5 levels of trait variability $* 10$ unique comparisons of trait aggregation methods $* 3$ scenarios). Overall, differences between the results of the aggregation methods for each simulated dataset increased with increasing trait variability. However, in most simulated datasets the different aggregation methods often resulted in similar trait affinities. Only 1.42 \% (213 out of 15.000) of all aggregated affinities showed a difference equal or greater than an absolute trait affinity of 0.1. Most of these differences occurred for the \textit{sim\_extreme} scenario (83.5 \%). The majority of the differences equal or above 0.1 were found between the aggregation methods \textit{direct\_agg\textsubscript{mean}} and \textit{stepwise\_agg\textsubscript{median}}, \textit{direct\_agg\textsubscript{median}} and \textit{stepwise\_agg\textsubscript{median}}, and \textit{stepwise\_agg\textsubscript{median}} and \textit{weighted\_agg} (Figure \ref{fig:sim_indv_runs}). 
% Differences were rather small, largest difference 0.265 for sim_extreme (stepwise_median vs weighted_agg)

@@

\subsection*{Re-analysis of Szöcs et al. using harmonised and aggregated grouping features}

@@

Using harmonised grouping features resulted in less traits that distinguish upstream and downstream sites in comparison to the original analysis (Figure \ref{fig:rda_species_scores} and Figure \ref{fig:boxplots_scores_on_constrained_axis}). According to the RDA of the trait composition, downstream sites were characterised by taxa with the traits multivoltinism, ovoviviparity and long life cycle duration. Upstream sites were characterised by univoltine taxa with a short life cycle duration that lay their eggs in an aquatic environment (aquatic eggs). The trait aquatic eggs of the harmonised grouping feature oviposition has been derived by amalgamating the trait oviposition in clutches and other related traits (Table \ref{tab:traits_harmonization}). The traits shredder and gill respiration did not characterise sites with high salinisation.

Using at family-level aggregated traits from the harmonised dataset showed also results similar to the original analysis but with fewer traits distinguishing upstream and downstream sites (Figure \ref{fig:rda_species_scores} and Figure \ref{fig:boxplots_scores_on_constrained_axis}). The \textit{direct\_agg\textsubscript{mean}}, \textit{direct\_agg\textsubscript{median}}, and \textit{weighted\_agg} 
characterised the downstream sites with the same traits as the original analysis except that downstream sites were not characterised by the trait shredder. Upstream sites were characterised by the traits univoltinism, aquatic eggs and short life cycle duration. The same results were obtained when re-analysing the data with traits aggregated by the \textit{stepwise\_agg\textsubscript{mean}} and \textit{stepwise\_agg\textsubscript{median}}, with the exception that none of the life cycle traits characterised upstream or downstream sites. Thus, in our comparison aggregating by \textit{direct\_agg\textsubscript{mean}}, \textit{direct\_agg\textsubscript{median}}, and \textit{weighted\_agg} yielded to the least change in interpretation of the RDA results compared to the original results.

\begin{figure}[H]
    \label{fig:rda_species_scores}
    \centering
    \includegraphics[width=16.5cm, height=10cm]{Species_scores_rda.png}
    \caption{Species scores obtained by RDA from the original analysis \cite{szocs_effects_2014}, using harmonised grouping features, and using harmonised grouping features with traits aggregated to family-level. The dots represent the individual species scores for each analysed trait along the conductivity axis. The violin plot shows the density estimate of the species scores. Gray vertical lines indicate the median of the obtained species scores.}
\end{figure}

@@

% Why do we need trait-based analysis, especially across regions?
Synthesizing trait information from multiple trait databases is crucial to unfold the full potential of trait-based approaches, e.g. for studying community trait responses to environmental gradients worldwide. Our attempt of harmonising invertebrate trait information from trait databases of different regions showed that grouping feature harmonisation is labour-intensive because grouping features are differentiated differently into traits between trait databases and various codings are used to describe traits. In addition, the same traits are sometimes defined differently in the investigated databases, requiring expert knowledge to reclassify these traits (e.g. the trait piercer in the Tachet database, Table \ref{tab:traits_harmonization}). Although, others have noted the lack of standardised trait terminology in freshwater ecology \cite{baird_toward_2011, brink_traits-based_2011} we are the first to describe invertebrate trait definition discrepancies for some commonly used grouping features.
%In addition we provide harmonised grouping feature trait datasets that can be used for other studies
To resolve definition discrepancies and facilitate data synthesis in the future terminological standards are needed. Harmonised definitions and concepts of traits have been developed in the past for other organism groups, e.g. for plants with the \textit{Thesaurus Of Plant characteristics} (TOP) initiative \cite{garnier_towards_2017}. The core of this harmonisation initiative is to provide standardised trait definitions and to draw connections to synonyms, related terms, and surrounding concepts by linking to other controlled vocabularies or ontologies in the field. Such initiatives could be a role model for freshwater ecologists to establish unambiguous terminologies for invertebrate traits. The existing freshwater invertebrate trait databases could be linked through standardised terminologies or ontologies, as suggested by Baird et al. \cite{baird_toward_2011} almost a decade ago. By following the recently proposed \textit{Ecological Trait-data Standard Vocabulary} (ETS) providers of invertebrate trait data could connect their traits to such ontologies \cite{schneider_towards_2019}. Once standardised trait definitions are established, these will improve trait data sharing, trait data processing when working with multiple trait databases, and ultimately interpretation of the derived results. % What is not mentioned here is how traits are measured, for which measurement protocols exist I believe. I think this is an separate issue.

@@

@@

Although comprehensive invertebrate trait databases have been developed for several regions in the past, data synthesis is difficult due to discrepancies in trait definitions. By providing an overview of these definition discrepancies we hope to set a starting point for the development of standardised trait terminology through which invertebrate trait databases can be linked. A consent on standard terminology and the subsequent development of ontologies are the next steps to facilitate trait-based analysis on large scales. As our analysis showed, some grouping features might need to be reclassified to fit into such a standardised terminology. We could show that trait affinities resulting from fuzzy coding and binary coding can be used together, but a uniform coding of traits is another problem that should be addressed in the course of trait standardisation.

With the increasing use of computer vision in ecology, species identification could become easier and more automated, and trait aggregation might be less needed \cite{milosevic_application_2020}. However, if traits need to be aggregated we showed that trait aggregation is often in conclusion with expert assignments, specifically aggregation approaches using the median. Which aggregation method the objectively better measure is cannot be answered, but when trait aggregation is necessary 1) highly variable traits and taxa where a few genera have a high number of species should be treated carefully, 2) coding of traits should be considered, and 3) weighting trait information based on the taxa present in the used trait database does not seem to have much influence. No matter which aggregation approach is chosen, for future studies in which researchers must aggregate traits, a measure of trait dispersion should be reported to indicate the uncertainty of the aggregated estimate.
